---
title: "Diabetes Among The Pima Indians: An Exploratory Analysis"
description: |
  In this post we will explore the Pima Indian dataset from the UCI repository. This post will aim to showcase different ways of thinking of your 
  data. Most novices to data science would rush into data preprocessing and not explore the data properly. The data cleaning stage can be subjective 
  at times and here I offer my own view and opinons on this dataset
author:
  - name: Asel Mendis
    url: https://www.linkedin.com/in/asel-mendis-a620399b/
date: 11-06-2018
output:
  radix::radix_article:
    self_contained: false
categories:
  - Data Preprocessing
  - Exploratory Analysis
  - R 
repository_url: https://github.com/aslm123/easydsrp
creative_commons: CC BY
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Data

The data used in the current project contains a number of diagnostic measures of type 2 diabetes in women of the Pima Indian heritage, and whether or not the individual has type 2 diabetes. The dataset was obtained from Kaggle at (https://www.kaggle.com/uciml/pima-indians-diabetes-database). There is a total of 768 observations and 9 variables. The variables in the dataset are:

  *	Pregnancies
    - Number of pregnancies. 
  
  *	Glucose
    - The blood plasma glucose concentration after a 2 hour oral glucose tolerance test.
  
  *	BloodPressure
    - Diastolic blood pressure (mm/HG).
  
  
  *	SkinThickness
    - Skinfold thickness of the triceps (mm).
  
  
  *	Insulin
    - 2 hour serum insulin (mu U/ml).
  
  
  *	BMI
    - Body mass index (kg/m squared)
  
  
  *	DiabetesPedigreeFunction
    - A function that determines the risk of type 2 diabetes based on family history, the larger the function, the higher the risk of type 2 diabetes.
  
  
  *	Age. 
    - Age (years)
  
  
  *	Outcome
    - Whether the person is diagnosed with type 2 diabetes (1 = yes, 0 = no).

## Preprocessing
```{r Libraries, message=FALSE}
library(readr)
library(tidyverse)
library(dplyr)
library(tidyr)
library(knitr)
library(pipeR)
```

```{r Working_Directory, message=FALSE, echo=FALSE, include=FALSE}
Diabetes <- read_csv("C:/Users/aselr/Documents/RMIT/SEMESTER 2 2018/Analysis of Categorical Data/Final Project/Phase 1/diabetes.csv") %>% 
  as.data.frame()
```


## Overview
```{r Overview}
glimpse(Diabetes)
```

```{r Summary}
mlr::summarizeColumns(Diabetes) %>% 
  kable(caption="Summary Table of Diabetes") 
```


## Type Conversions

### Factors

The outcome variable needs to be converted into a categorical variable. It will be ordered to accomodate analysis purposes in the near future.

```{r}
Diabetes$Outcome <- as.factor(unlist(Diabetes$Outcome))
Diabetes$Outcome <- factor(Diabetes$Outcome, levels=c("1", "0"),
                           labels = c("Positive", "Negative"))
summary(Diabetes$Outcome)
```
We can see that there are almost twice as many people wihtout diabetes than there are with diabetes. While there is no universal cutoff for the number of rows for your target variable, this should suffice. 

*We just need to make sure that there is enough data for your model to learn how to differentiate between the two.*

### Numeric

The following variables:

	* Glucose

	* BloodPressure

	* SkinThickness

	* Insulin

	* BMI 

need to be converted into numeric variables from its current integer class. These variables have decimal values and its absence could potentially create misleading results and distort specified ranges of risk represented by those medical tests.

```{r}
Diabetes$Glucose <-  as.numeric(Diabetes$Glucose)
Diabetes$BloodPressure <-  as.numeric(Diabetes$BloodPressure)
Diabetes$SkinThickness <-  as.numeric(Diabetes$SkinThickness)
Diabetes$Insulin <-  as.numeric(Diabetes$Insulin)
Diabetes$BMI <-  as.numeric(Diabetes$BMI)
Diabetes$Age <- as.integer(Diabetes$Age)
```


## Dealing with Missing Values
${6/9}$ variables in the dataset have a number of zero markers. It appears that after taking the sum of each column and row separately, there are 763 `0` values in the dataset. Alarmingly, this represents almost 100% of our observations.

```{r}
list(
`Column` = colSums(Diabetes==0),
Row = sum(rowSums(Diabetes==0))
)
```

### Pregnancies
A value of '0' does not necessarily mean it is a missing value. For example:
*A woman has a zero record of pregancies because that woman has not been pregnant.*

This is one example on how careful you have to be when preprocessing your data for missing values. Zero does not mean it is missing. However in addition to that point, zero values could create problems when passing it to a machine learning model. Techniques like Regression can give you an estimate of the output when the variable is zero. Another method of dealing with zeros could be to bin the variable which will then create a categorical variable.

```{r}
Diabetes$Pregnancies <- ifelse(Diabetes$Pregnancies==0, "No", "Yes") %>% 
                            factor()
colSums(Diabetes==0)
summary(Diabetes$Pregnancies)
```

### Insulin

50% of the rows in Insulin have 0 values. Lets look at it in a practical sense. There are times when the body produces little to no insulin which is a sign on Type 1 Diabetes. But I doubt this is the case in this instance. Insulin is such an important variable concerning Diabetes, but when a variable is just rife with missing values you have to do something about it. You can just impute the values, but this is medical data and half of the values are missing. It would not be appropriate to just impute the rows with it mean in my opininon. That's why no matter how important it is, it has to be **REMOVED**. This is one of those times where you need to make a tough call.

```{r}
Diabetes$Insulin <- NULL
colSums(Diabetes==0)
```

### Skin Thickness

The same can be said for `SkinThickness`. 31% of its rows have 0 as a value. Practically how can a human have 0mm skinfold thickness. This variable is also not giving us much use. Therefore I will remove this variable as well.

```{r}
Diabetes$SkinThickness <- NULL
colSums(Diabetes==0)
```
 

### BMI
BMI can be supplemented with its respective range of obesity criteria. BMI only has 11 0 values so it would not be expected to cause too much trouble. If it was substantially higher, binning would not really solve the problem because the allocated bin may not be its correct one.

```{r}
Diabetes$BMI <- 
        ifelse(Diabetes$BMI<19,"Underweight",
				ifelse(Diabetes$BMI>=19 & 
				         Diabetes$BMI<=25, "Normal", 
				ifelse(Diabetes$BMI>=25 & 
				  Diabetes$BMI<=30, "Overweight","Obese"))) %>% 
    factor(levels=c("Underweight","Normal",
                "Overweight","Obese"))

list(BMI = summary(Diabetes$BMI))

```

So now I see that `Underweight` has 15 rows which means there were 15 rows that had a BMI less than 19 and knowing that `BMI` had 11 zero values, we know that most of the rows are zero. Because it is so little, I am choosing to keep it. Also, we have already removed two variables and I want to retain as much data as possible while making sure it can add value.



### Glucose

With a 2 hour 'Oral Glucose Tolerance Test' (OGTT), people are considered to not have diabetes if after 2 hours of administering the test their levels are below 7.8 mmol/L.
To categorize the levels of glucose tolerance, we will use the following criteria:

	* Hypoglycemia (Low Blood Sugar) - <2.2 mmol/L

	* Normal/No Diabetes - >=2.2 mmol/L - <=7.8mmol/L

	* Prediabetes (Hyperglycemia / High Blood Sugar) - >7.8 mmol/L - <=11.1 mmol/L

	* Diabetes - >11.1 mmol/L

**Although one of the levels says this person has diabetes, it is not a final diganosis. Other factors will have to be taken to account. To avoid misleads, 'Diabetes' in this circumstance REFERS TO THE TEST RESULT ANME AND NOT A FINAL DIAGNOSIS OF TYPE 2 DIABETES'**

The unit of measurement for the 2-hour OGTT in this dataset is assumed to be in milligrams per deciliter (mg/dl). It can be converted to Milimoles per liter (mmol/l) so that we may appply a qualitative test result to the numeric results. Multiplying the current results by 0.0555 will convert them to be measured in mmol/l.

```{r}
Diabetes$Glucose <- Diabetes$Glucose*0.0555
```

```{r}
Diabetes$Glucose <-
	if_else(Diabetes$Glucose<2.2,"Hypoglycemia",
	if_else(Diabetes$Glucose>=2.2 & 
	          Diabetes$Glucose<=7.8,"Normal",
	if_else(Diabetes$Glucose>7.8 & 
	          Diabetes$Glucose<=11.1,
	        "Hyperglycemia","Diabetes"))) %>% 
  factor()
```

```{r}
list(
`Test Result` = summary(Diabetes$Glucose)
)
```

It appears that 74% of participants have Normal Glucose levels and 25% have Prediabetes/Hyperglycemia/High Blood Sugar.Only 1% have Hypoglycemia/Low Blood Sugar.

My main concern with Hypoglycemia is that when I take it through a machine learning process, the level may not be present during k-fold cross validation. I cannot say for sure at this point but it is worth noting and keeping in mind for the future.


## Blood Pressure
When measuring Blood Pressure, two measures are used:

  	* Systolic
    	- Measures the pressure in blood vessels when the heart beats. 

  	* Diastolic
    	- Measures the pressure in blood vessels when the heart rests between beats. 
    
In this dataset, only the diastolic blood Pressure is reported. This requires us to proceed with caution. 

As reported by the Centers for Disease Control and Prevention, blood pressure is measured using systolic and diastolic units. When they are reported we can make the following inferences:

  * Normal
  	- Systolic <120mmHg,
  	- Diastolic <80mmHg   

  * Prehypertension 
    - Systolic >=120mmHg - <=139mmHg
    - Diastolic >=80 - <=89mmHg 

  * High 
 	  - Systolic: >=140mmHg
  	- Diastolic: >=90mmHg 

BECAUSE SYSTOLIC PRESSURE IS NOT REPORTED, I DO NOT FEEL COMFORTABLE IN PROCEEDING WITH THIS VARIBALE. THIS IS A SUBJECTIVE DECISION OF MINE.

```{r}
Diabetes$BloodPressure <- NULL
```


```{r echo=FALSE}
load('C:/Users/aselr/Documents/R Programming/mode.rda') # Function to calculate mode
```

```{r}
str(mode)
```

# Visualisation of variables 

## Numeric Variables

### Age
```{r}
ggplot(Diabetes, aes(y=Age, x=Outcome)) + 
      geom_boxplot() + geom_jitter()+
      theme_bw() +
      xlab("Outcome") + ylab("Age") +
      stat_summary(fun.y=mode, colour="Orange", 
                  geom="point", shape=16, size=5) +
      stat_summary(fun.y=mean, colour="purple", 
                  geom="point", shape=16, size=5) +
      ggtitle(label="Age by Outcome",
              subtitle = "Orange=Most Frequent\nPurple=Average Age") +
      theme(axis.text.x = element_text(face="bold",size=12),
            axis.text.y = element_text(face="bold",size=12),
            title = element_text(face="bold",size=12),
            
            axis.title = element_text(face="bold",size=12)) +
      scale_y_continuous(breaks=seq(20,80,4))
```

It seems there is a lot of noise in this variable. There is no clear distinction in age with regards to having diabetes.




### Diabetes Pedigree Function
```{r}
ggplot(Diabetes, aes(y=DiabetesPedigreeFunction, x=Outcome)) + 
      geom_boxplot() + geom_jitter()+
      theme_bw() +
      xlab("Outcome") + ylab("DiabetesPedigreeFunction") +
      stat_summary(fun.y=mode, colour="orange", 
                  geom="point", shape=16, size=5) +
      stat_summary(fun.y=mean, colour="purple", 
                  geom="point", shape=16, size=5) +
      ggtitle(label="Diabetes Pedigree Function by Outcome",
              subtitle = "Orange=Most Frequent\nPurple=Average Age") +
      theme(axis.text.x = element_text(face="bold",size=12),
            axis.text.y = element_text(face="bold",size=12),
            title = element_text(face="bold",size=12),
            axis.title = element_text(face="bold",size=12)) +
      scale_y_continuous(breaks=seq(0,3,0.5))
```

Interestingly, the `Diabetes Pedigree Function` does not seem to give a clear picture of a diabetic outcome. This is supposed to be a score wherein the higher the score, the more likely you are to have diabetes. This is also a variable with a lot of noise. 


## Categorical Variables (% of Outcome) 

### Pregnancies

```{r}
(pregnant <- table(Diabetes$Pregnancies, Diabetes$Outcome,
                          dnn = c("Pregnant", "Outcome")) )
pregnant %>% prop.table(2) %>% round(2) %>% 
  kable(format = 'html')
```

It seems that having a pregnancy does not necessarily increase your chances of having diabetes as the same proportion of women who had or didnt have diabetes had at least one pregnancy.


### Obesity
```{r}
(bmi <- table(Diabetes$BMI, Diabetes$Outcome,
                          dnn = c("BMI", "Outcome"))  )
bmi %>% prop.table(2)%>% round(2) %>% 
  kable(format = 'html')
```

Unsurprisingly, 80% of Diabetic women were obese while 16% were overweight. Only 3% were reported to be of normal weight. Among the women that do not have diabetes, 50% were obese, 27% overweight and 20% normal. 

### Glucose
```{r}
(glucose <- table(Diabetes$Glucose, Diabetes$Outcome,
                          dnn = c("Glucose Level", "Outcome")) )
glucose %>% prop.table(2) %>% round(2) %>% 
  kable(format = 'html')
```

49% of women who have diabetes were positive for Hyperglycemia and 50% had normal glucose levels. Surprisingly, the glucose levels do not seem to clearly differentiate between those who are diabetic. Obviously, people with Hyperglycemia are more likely to have diabetes but the magnitude is very low according to the above table.

Unsurprisingly, 87% of women without diabetes had normal glucose levels.


# Final Data


```{r}
summary(Diabetes)
```






